Missing the Point: Convergence of Multiple Imputation by Chained Equations Algorithms

### 2020-03-22 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Comments and changes to research report manuscript to fit new simulation (until methods section)

  pdftitle={Missing the Point: Convergence of Multiple Imputation using Chained Equations Algorithms},

@@

\title{Missing the Point: Convergence of Multiple Imputation using Chained
Equations Algorithms}
\author{Hanne I. Oberman}
\date{16-3-2020}

@@

At some point, any scientist conducting statistical analyses will run
into a missing data problem (Allison 2001). Missingness is problematic
because statistical inference cannot be performed on incomplete data
without employing \emph{ad hoc} solutions (e.g., list-wise deletion),
which may yield wildly invalid results (Van Buuren 2018). A popular
answer to the ubiquitous problem of missing information is to use the
framework of multiple imputation (MI), proposed by Rubin (1987). For the
theoretical foundation of MI, see Rubin (1987). For an accessible and
comprehensive introduction to MI from an applied perspective, see
e.g.~Van Buuren (2018). With MI, missing datapoints are `imputed' (i.e.,
filled in) via an \emph{imputation model}. The imputation model can be a
joint model for all variables, or several univariate conditional
distributions (fully conditional specification). FCS is more flexible
and ``avoids the problem of specifying an appropriate joint imputation
distribution and replaces this by the selection of appropriate
univariate conditional distributions''. With FCS, an iterative
algorithmic procedure is used to \ldots{} multiple imputation using
chained equations \ldots{} ``The values for imputing the missing data
are drawn from the posterior predictive distribution of the missing data
conditional on the observed data.'' There is no scientific consensus on
the convergence properties of FCS algorithms (Takahashi 2017). Some
default techniques in \texttt{mice} might not yield converged states at
all (Murray 2018). Therefore, algorithmic convergence should be
monitored carefully.

Current practice is visually inspecting imputation chains for signs of
non-convergence, but may be challenging to the untrained eye (Van Buuren
2018, \textasciitilde6.5.2). Therefore diagnostic evaluation of
convergence would be preferred. However, not known whether MCMC
diagnostics are appropriate for MICE \ldots{} The application of
convergence diagnostics to MI methods has not been systematically
studied (Van Buuren 2018). How can non-convergence be diagnosed? How
many iterations are sufficient/required? What are the effects of
non-convergence on estimates, predictions and inference? What is an
appropriate non-convergence diagnostic? Are the default number of
iterations sufficient?

@@

Defaults is: SPSS (10;
\url{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}),
Mplus (100??;
\url{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}),
Stata (10; \url{https://www.stata.com/manuals13/mi.pdf}, p.~139), Amelia
(NA, because resampling, not convergence;
\url{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}), en MI
(30; \url{https://cran.r-project.org/web/packages/mi/mi.pdf}).

@@

\hypertarget{terminology}{%
\subsection{Terminology}\label{terminology}}

Let \(Y\) denote an \(n \times p\) matrix containing the data values on
\(p\) variables for all \(n\) units in a sample. The collection of
observed data values in \(Y\) is denoted by \(Y_{obs}\), and will be
referred to as `incomplete' or `observed' data. The missing part of
\(Y\) is denoted by \(Y_{mis}\). Which parts of \(Y\) are missing is
determined by the `missingness mechanism'.\\
This note only considers a `missing completely at random' (MCAR)
mechanism, where the probability of being missing is equal for all
\(n \times p\) cells in \(Y\) (Rubin 1987).

Figure 1 provides an overview of the steps involved with MI---from
incomplete data, to \(m\) completed datasets, to \(m\) estimated
quantities of interest (\(\hat{Q}\)s), to a single pooled estimate
\(\bar{Q}\). This note focuses on the algorithmic properties of the
imputation step.

The MI algorithm in \texttt{mice} has an iterative nature. For each
missing datapoint in \(Y_{mis}\), \(m\) `chains' of potential values are
sampled. Only the ultimate sample that each chain lands on is imputed.
The number of iterations per chain will be denoted with \(T\), where
\(t\) varies over the integers \(1, 2, \dots, T\). This is repeated
\(m\) times. Imputed values are the ultimate samples of a `chain' of For
each imputed value (\(t = T\)), a s for each missing datapoint in
\(Y_{mis}\), . Each of the \(m\) chains starts with an initial value,
drawn randomly from \(Y_{obs}\). The chains are terminated after a
predefined number of iterations. , and subsequently used in the analysis
and pooling steps. The collection of samples between the initial value
(at \(t=1\)) and the imputed value (at \(t=T\)) will be referred to as
an `imputation chain'.

@@

In iterative algorithmic procedures (Markov chain Monte Carlo methods;
e.g., \texttt{mice} algorithms or Gibbs samplers) non-convergence may be
present as non-stationarity within chains (i.e., trending), or as slow
mixing between chains (i.e., no intermingling).

@@

The mixing component can be assessed with the potential scale reduction
factor \(\widehat{R}\) (a.k.a. `Gelman-Rubin statistic'; Gelman and
Rubin 1992). With an adapted version of \(\widehat{R}\), proposed by
Vehtari et al. (2019), we might also evaluate the stationarity component
of convergence. This would make \(\widehat{R}\) a general convergence
diagnostic. The application of \(\widehat{R}\) to assess stationarity
has not been thoroughly investigated. Therefore, this study employs both
\(\widehat{R}\) and autocorrelation to investigate convergence, as
recommended by (Cowles and Carlin 1996, \textasciitilde898).

The mixing component of convergence may be evaluated with the potential
scale reduction factor, \(\widehat{R}\) (a.k.a. `Gelman-Rubin
statistic'; Gelman and Rubin 1992). The stationarity component may be
assessed through autocorrelations. An adapted version of \(\widehat{R}\)
might also be used to diagnose non-stationarity. Other convergence
diagnostics are outside of the scope of this study.

All of these methods evaluate the convergence of univariate scalar
summaries (e.g., chain means or variances). These convergence
diagnostics cannot diagnose convergence of multivariable statistics
(i.e., relations between scalar summaries). Van Buuren (2018) proposed
to implement multivariable evaluation through eigenvalue decomposition
(MacKay and Mac Kay 2003). This method is outside of the scope of the
current study (\textbf{not anymore!}).

To define \(\widehat{R}\), we follow notation by (Vehtari et al. 2019,
\textasciitilde5). Let \(M\) be the total number of chains, \(T\) the
number of iterations per chain, and \(\theta\) the scalar summary of
interest (e.g., chain mean or chain variance). For each chain
(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
average these to obtain within-chain variance \(W\).

@@

Following the same notation, we define autocorrelation as the
correlation between two subsequent \(\theta\)-values within the same
chain (Lynch 2007, \textasciitilde147). In this study we only consider
\(AC\) at lag 1, i.e., the correlation between the \(t^{th}\) and
\(t+1^{th}\) iteration of the same chain.

@@

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods (i.e., average bias, average confidence interval
width, and empirical coverage rate across simulations; Van Buuren 2018,
\textasciitilde2.5.2). That is, there is no baseline measure available
to evaluate performance against.The current practice of visually
inspecting imputation chains for signs of non-mixing or non-stationarity
can only diagnose severely pathological cases of non-convergence. We
could also look at distributional characteristics, and plausibility of
imputed values, see Vink (n.d.). For now, this is outside of the scope
of this study.

Based on an empirical finding (Lacerda, Ardington, and Leibbrandt 2007),
we hypothesize that \(\widehat{R}\) will over-estimate non-convergence
of MI algorithms. The threshold of \(\widehat{R} < 1.01\) will then be
too stringent for diagnosing convergence. This over-estimation may,
however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed (Brooks and Gelman 1998, p\textasciitilde437). In
\texttt{mice}, initial values are chosen randomly from the observed
data. Therefore, we cannot be certain that the initial values are
over-dispersed. We expect this to have little effect on the hypothesized
performance of \(\widehat{R}\). No hypothesis was formulated about the
performance of \(AC\) as convergence diagnostic.

@@

We use model-based simulation in \texttt{R} to perform multiple
imputation under 100 simulation conditions. In each condition, the MI
algorithm comprises of a different imputation chain length
(\(T = 1, 2, \dots, 100\)). The number of simulation runs per condition
is 1000. For each simulation condition (i.e., number of iterations) and
each repetition (i.e., simulation run) we compute convergence
diagnostics (\(\widehat{R}\) and autocorrelation) and simulation
diagnostics (bias, confidence interval width and coverage). These
diagnostics are aggregated across repetitions to obtain their average
value per simulation condition. The simulation set-up is summarized in
the pseudo-code below. The complete \texttt{R} script of the simulation
study is available from
\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

@@

\hypertarget{data-simulation}{%
\subsubsection{Data Simulation}\label{data-simulation}}

The point of origin for all simulation conditions is a finite population
of \(N=1000\). The data are simulated to solve a multiple linear
regression problem, where dependent variable \(Y\) is regressed on
independent variables \(X_1\), \(X_2\) and \(X_3\):

@@

The quantity of scientific interest is regression coefficient
\(\beta_1\). The data generating model is a multivariate normal
distribution with the following means structure and variance-covariance
matrix:

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3\\
\epsilon
\end{pmatrix} \sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\\[2\jot].
\end{align*}

Data is simulated with the function \texttt{mvtnorm::rmvnorm()}, and
outcome variable \(Y\) is subsequently calculated as
\[Y =  2X_1 + .5X_2 - X_3 + \epsilon .\] The complete data is `amputed'
once for each simulation repetition. That is, the function
\texttt{mice::ampute()} is used to impose a missingness mechanism upon
the data. The missingness is univariate, and the probability to be
missing is the same for all four variables, namely 20\%
(\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This leaves 20\% of the rows
completely observed. \%The resulting amputed data is equal for all
simulation conditions in the same repetition.

@@

From the imputation step, we extract chain means and chain variances
across iterations (\(\theta\) at \(t = 1,2,\dots,T\)).\\
We compute convergence diagnostics separately for both \(\theta\)s, and
for each of the four variables in the dataset. We report the maximum
(absolute) value as the `worst linear predictor' of convergence.
\(\widehat{R}\) is computed by implementing Vehtari et al.'s (2019)
recommendations. \(AC\) is computed with function \texttt{stats::acf()}.

@@

\begin{itemize}
\item
  Remove the table.
\item
  Add more info about figure legends and axes.
\end{itemize}

Figures 2 and 3 display results per simulation condition
(\(T = 1,2,\dots,100\)).

\hypertarget{convergence-diagnostics-1}{%
\subsubsection{Convergence
Diagnostics}\label{convergence-diagnostics-1}}

@@

\hypertarget{simulation-diagnostics}{%
\subsubsection{Simulation Diagnostics}\label{simulation-diagnostics}}

>>>>

  pdftitle={Missing the Point: Convergence of Multiple Imputation by Chained Equations Algorithms},

@@

\title{Missing the Point: Convergence of Multiple Imputation by Chained
Equations Algorithms}
\author{Hanne I. Oberman}
\date{16-3-2020}

@@

At some point, any scientist conducting statistical analyses will run
into a missing data problem (Allison 2001). Missingness is problematic
because statistical inference cannot be performed on incomplete data
without employing \emph{ad hoc} solutions (e.g., list-wise deletion),
which may yield wildly invalid results (Van Buuren 2018). A popular
answer to the ubiquitous problem of missing information is to use the
framework of multiple imputation (MI), proposed by Rubin (1987). For the
theoretical foundation of MI, see Rubin (1987). For an accessible and
comprehensive introduction to MI from an applied perspective, see
e.g.~Van Buuren (2018). With MI, missing datapoints are `imputed' (i.e.,
filled in) via an \emph{imputation model}. The imputation model can be a
joint model for all variables, or several univariate conditional
distributions (fully conditional specification). FCS is more flexible
and ``avoids the problem of specifying an appropriate joint imputation
distribution and replaces this by the selection of appropriate
univariate conditional distributions''. With FCS, an iterative
algorithmic procedure is used to \ldots{} multiple imputation by chained
equations \ldots{} ``The values for imputing the missing data are drawn
from the posterior predictive distribution of the missing data
conditional on the observed data.'' There is no scientific consensus on
the convergence properties of FCS algorithms (Takahashi 2017). Some
default techniques in \texttt{mice} might not yield converged states at
all (Murray 2018). Therefore, algorithmic convergence should be
monitored carefully.

@@

Current practice is visually inspecting imputation chains for signs of
non-convergence, but may be challenging to the untrained eye (Van Buuren
2018, \(\S\) 6.5.2). Therefore diagnostic evaluation of convergence
would be preferred. However, not known whether MCMC diagnostics are
appropriate for MICE \ldots{} The application of convergence diagnostics
to MI methods has not been systematically studied (Van Buuren 2018). How
can non-convergence be diagnosed? How many iterations are
sufficient/required? What are the effects of non-convergence on
estimates, predictions and inference? What is an appropriate
non-convergence diagnostic? Are the default number of iterations
sufficient? Defaults in: SPSS = 10
\href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
Mplus = 100??
\href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
Stata = 1;
\href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
Amelia = NA, because resampling, not convergence
\href{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}{(link)},
en MI = 30
\href{https://cran.r-project.org/web/packages/mi/mi.pdf}{(link)}.

@@

\hypertarget{notation}{%
\subsection{Notation}\label{notation}}

Let \(y\) denote an \(n \times p\) matrix containing the data values on
\(p\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, \dots, n\)) on variable \(j\) (\(j = 1, \dots, p\)) may
be either observed or missing. The collection of observed data values in
\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
to as \(y_{mis}\). For each datapoint in \(y_{mis}\) \(M \times T\)
times plausible values are sampled, where \(M\) is the number of
imputations (\(m = 1, \dots, M\)) and \(T\) is the number of iterations
(\(t = 1, \dots, T\)). The collection of samples between the initial
value (at \(t=1\)) and the imputed value (at \(t=T\)) will be referred
to as an `imputation chain'.

@@

In iterative algorithmic procedures (Markov chain Monte Carlo methods;
e.g., \texttt{mice} algorithms or Gibbs samplers) non-convergence may be
present as non-stationarity within chains (i.e., trending), or as slow
mixing between chains (i.e., no intermingling). (\textbf{add
example/guidelines visual inspection here?})

@@

The mixing component can be assessed with the potential scale reduction
factor \(\widehat{R}\) (a.k.a. `Gelman-Rubin statistic'; Gelman and
Rubin 1992). With an adapted version of \(\widehat{R}\), proposed by
Vehtari et al. (2019), we might also evaluate the stationarity component
of convergence. This would make \(\widehat{R}\) a general convergence
diagnostic. The application of \(\widehat{R}\) to assess stationarity
has not been thoroughly investigated. Therefore, this study employs both
\(\widehat{R}\) and autocorrelation to investigate convergence, as
recommended by (Cowles and Carlin 1996, 898).

Note that all of these methods evaluate the convergence of univariate
scalar summaries (e.g., chain means or variances). These convergence
diagnostics cannot diagnose convergence of multivariable statistics
(i.e., relations between scalar summaries). Van Buuren (2018) proposed
to implement multivariable evaluation through eigenvalue decomposition
(MacKay and Mac Kay 2003). This method is outside of the scope of the
current study (\textbf{not anymore!}).

\hypertarget{potential-scale-reduction-factor}{%
\subsection{Potential scale reduction
factor}\label{potential-scale-reduction-factor}}

To define \(\widehat{R}\), we follow notation by (Vehtari et al. 2019,
5). Let \(M\) be the total number of chains, \(T\) the number of
iterations per chain, and \(\theta\) the scalar summary of interest
(e.g., chain mean or chain variance). For each chain
(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
average these to obtain within-chain variance \(W\).

@@

\hypertarget{autocorrelation}{%
\subsection{Autocorrelation}\label{autocorrelation}}

Following the same notation, we define autocorrelation as the
correlation between two subsequent \(\theta\)-values within the same
chain (Lynch 2007, 147). In this study we only consider \(AC\) at lag 1,
i.e., the correlation between the \(t^{th}\) and \(t+1^{th}\) iteration
of the same chain.

@@

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods (i.e., average bias, average confidence interval
width, and empirical coverage rate across simulations; Van Buuren 2018,
\(\S\) 2.5.2). That is, there is no baseline measure available to
evaluate performance against.The current practice of visually inspecting
imputation chains for signs of non-mixing or non-stationarity can only
diagnose severely pathological cases of non-convergence. We could also
look at distributional characteristics, and plausibility of imputed
values, see Vink (n.d.). For now, this is outside of the scope of this
study.

Based on an empirical finding (Lacerda, Ardington, and Leibbrandt 2007),
we hypothesize that \(\widehat{R}\) will over-estimate non-convergence
of MI algorithms. The threshold of \(\widehat{R} < 1.01\) will then be
too stringent for diagnosing convergence. This over-estimation may,
however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed (Brooks and Gelman 1998, 437). In \texttt{mice}, initial
values are chosen randomly from the observed data. Therefore, we cannot
be certain that the initial values are over-dispersed. We expect this to
have little effect on the hypothesized performance of \(\widehat{R}\).
No hypothesis was formulated about the performance of \(AC\) as
convergence diagnostic.

@@

The convergence of \texttt{mice} is investigated through model-based
simulation in \texttt{R} (version 3.6.3; R Core Team 2019). The
simulation set-up is summarized in the pseudo-code below. The complete
\texttt{R} script of the simulation study is available from
\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

@@

A finite population of \(N=1000\) is simulated to solve a multiple
linear regression problem, where dependent variable \(Y\) is regressed
on independent variables \(X_1\), \(X_2\) and \(X_3\):

@@

The data generating model is a multivariate normal distribution

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3\\
\epsilon
\end{pmatrix} \sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\\[2\jot]
\end{align*}

and simulated with the function \texttt{mvtnorm::rmvnorm()}. Outcome
variable \(Y\) is subsequently calculated as
\[Y =  2X_1 + .5X_2 - X_3 + \epsilon .\]

The complete data is `amputed' once for each simulation repetition. That
is, the function \texttt{mice::ampute()} is used to impose a missingness
mechanism upon the data. \textbf{(change this: The missingness is
univariate, and the probability to be missing is the same for all four
variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This
leaves 20\% of the rows completely observed). }

@@

\textbf{(tot hier doorgelezen)}

\(\widehat{R}\) is computed by implementing Vehtari et al.'s (2019)
recommendations. \(AC\) is computed with function \texttt{stats::acf()}.

@@

\begin{itemize}
\tightlist
\item
  Add more info about figure legends and axes.
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# plot results}
\KeywordTok{load}\NormalTok{(}\StringTok{"../Results/results.Rdata"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{convergence-diagnostics-1}{%
\subsection{Convergence Diagnostics}\label{convergence-diagnostics-1}}

@@

\hypertarget{simulation-diagnostics}{%
\subsection{Simulation Diagnostics}\label{simulation-diagnostics}}
